Collaborative Multioutput Gaussian Processes: \\ Supplementary Material

### 2014-03-19 paper/inference.tex
3.2 finished

So far in the description of the model and inference, it is implicitly assumed that every output has full observations at all of the inputs $\X$.
To discern where learning occurs for each output, we make the missing data scenario more explicit.
Specifically, each output $i$ can have observations at a different set of inputs $\X_i$. 
We shall use $\oi$ to denote the indice of $\X_i$ (in the set $\X$) and use the indexing operator $\B(\oi)$ to select the rows corresponding to $\oi$ from any arbitrary matrix $\B$.
We also overload $\y_i$ as the observed targets of output $i$.

%------------------------------------------
% derivatives of q(u_j)
\newcommand{\Lgj}{\calL^g_j}
\newcommand{\ynoj}{\y_i^{\backslash j}}
\newcommand{\Kjzz}{\mat{K}_{jzz}}
From (eq. \ref{eq:L}), we can obtain the derivatives of the ELBO wrt the variational parameters for optimization.
The derivatives of $\calL$ wrt the parameters of $q(\u_j)$ are given by:
\begin{align}
\deriv{\calL}{\m_j} 
=& \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj \\
\nonumber
&- \bigg[\Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi) \bigg] \m_j \text{,} \\
\deriv{\calL}{\S_j} 
=& \frac{1}{2} \S_j^{-1} - \frac{1}{2} \bigg[ \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi) \bigg],
\end{align}
where $\y_i^{\backslash j} = \y_i - \Ahi(\oi) \mhi - \sum_{j' \neq j} w_{ij'} \A_{j'}(\oi) \m_{j'}$.
%\noindent As a function of the parameters of $q(\u_j)$, the lowerbound $\calL$ is:
%\begin{align}
%\nonumber
%\calL^g_j \define&
% \sum_{i=1}^P \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)  \\
% \nonumber
% &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_{ij}^2 \tilde{\K}_j(\oi,\oi) 
% + \beta_i \trace w_{ij}^2 \S_j \A_j(\oi,:)^T \A_j(\oi,:) \bigg)
% \\
% &- \frac{1}{2} \log |k(\Zj,\Zj) \S_j^{-1}| -\frac{1}{2} \trace k(\Zj,\Zj)^{-1} (\m_j \m_j^T + \S_j) ,
%\end{align}
%where $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$, which gives $\A_j(\oi,:) = k(\X_i,\Zj) k(\Zj,\Zj)^{-1}$, and  
%$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$. \\

@@

Since the optimal distributions $q(\u_j)$ and $q(\v_i)$ are in the exponential family, it is more convenient to use stochastic variational inference \citep{hensman2012fast,hensmangaussian} to perform update of their canonical parameters.
This works by taking a step of length $l$ in the direction of the natural gradient approximated by mini-batches of the data.
For instance, consider $q(\u_j)$ whose canonical parameters are $\Phi_1 = \S_j^{-1} \m_j$ and $\Phi_2 = -\frac{1}{2}\S_j^{-1}$.
Their stochastic update equations at time $t + 1$ are given by:
\begin{align}
\Phi_{1(t+1)} &= \S_{j(t)}^{-1} \m_{j(t)} + l \bigg(\sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj - \S_{j(t)}^{-1} \m_{j(t)} \bigg) \\
\Phi_{2(t+1)} &= -\frac{1}{2} \S_{j(t)}^{-1} + l \bigg(\frac{1}{2} \S_{j(t)}^{-1} - \frac{1}{2} \mat{\Lambda} \bigg),
\end{align}
\normalsize
where $\mat{\Lambda} = \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi)$.

%TODO table for complexity
\subsection{INDUCING INPUTS AND HYPER-PARAMETERS \label{sec:hyperparameters}}
To learn the hyperparameters, which in this model include the mixing weights, the covariance hyperparameters of the latent processes, and the noise precision of each output, we follow standard practice in GP inference.
For this model, this involves taking derivatives of the ELBO and applying standard stochastic gradient descent in alternative steps with the variational parameters, much like a variational EM algorithm. 
The derivatives are given in the supplementary material.

Learning of the inducing inputs, which was not considered in the single output case in \citep{hensmangaussian}, is also possible in our stochastic optimization approach.
In the supplementary material, we show that with the help of vectorized operations, the derivatives of the lower bound w.r.t the inducing inputs can be computed with a cost \emph{independent} of the input dimension.
This makes optimizing the inducing locations a practical option, which can be important for e.g. in high-dimensional problems where fixing the inducing inputs may hinder the performance.
Indeed, our experiments on a large scale multioutput problem shows that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation.

\subsection{COMPLEXITY ANALYSIS}
In this section we analyze the complexity of the model and compare it to existing multioutput approaches.
For consistency, we first unify common notations used for all models.
We use $P$ as the number of outputs, $N$ as the number of inputs, $Q$ as the number of shared latent processes, $M_t$ as the \emph{total} number of inducing inputs.
It is worth noting that $M_t = (P + Q) \times M$ in our COGP model, assuming that each sparse process has equal number of inducing points. 
Also, COGP have $P$ additional individual processes, one for each output.

>>>>

So far in the description of the model and inference method we have
 implicitly assumed that every output has full observations at all inputs $\X$.
To discern where learning occurs for each output, we make the missing data scenario more explicit.
Specifically, each output $i$ can have observations at a different set of inputs $\X_i$. 
We shall use $\oi$ to denote the indices of $\X_i$ (in the set $\X$) and use the indexing operator $\B(\oi)$ to select the rows corresponding to $\oi$ from any arbitrary matrix $\B$.
We also overload $\y_i$ as the observed targets of output $i$. 
%
\subsubsection{Learning the Parameters of the Variational Distribution}
%------------------------------------------
% derivatives of q(u_j)
\newcommand{\Lgj}{\calL^g_j}
\newcommand{\ynoj}{\y_i^{\backslash j}}
\newcommand{\Kjzz}{\mat{K}_{jzz}}
 We can obtain the derivatives of the ELBO in Equation  \eqref{eq:L}
  wrt the variational parameters for optimization.
The derivatives of $\calL$ wrt the parameters of $q(\u_j)$ are given by:
\begin{align}
\deriv{\calL}{\m_j} 
=& \sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj \\
\nonumber
&- \bigg[\Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi) \bigg] \m_j \text{,} \\
\deriv{\calL}{\S_j} 
=& \frac{1}{2} \S_j^{-1} - \frac{1}{2} \bigg[ \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi) \bigg],
\end{align}
where $\y_i^{\backslash j} = \y_i - \Ahi(\oi) \mhi - \sum_{j' \neq j} w_{ij'} \A_{j'}(\oi) \m_{j'}$.
%\noindent As a function of the parameters of $q(\u_j)$, the lowerbound $\calL$ is:
%\begin{align}
%\nonumber
%\calL^g_j \define&
% \sum_{i=1}^P \log \Normal(\y_i; \sum_{j=1}^Q w_{ij} \A_j(\oi,:) \m_j + \Ahi \mhi, \beta_i^{-1} \I)  \\
% \nonumber
% &- \frac{1}{2} \sum_{i=1}^P \bigg(\beta_i \trace w_{ij}^2 \tilde{\K}_j(\oi,\oi) 
% + \beta_i \trace w_{ij}^2 \S_j \A_j(\oi,:)^T \A_j(\oi,:) \bigg)
% \\
% &- \frac{1}{2} \log |k(\Zj,\Zj) \S_j^{-1}| -\frac{1}{2} \trace k(\Zj,\Zj)^{-1} (\m_j \m_j^T + \S_j) ,
%\end{align}
%where $\A_j = k(\X,\Zj)k(\Zj,\Zj)^{-1}$, which gives $\A_j(\oi,:) = k(\X_i,\Zj) k(\Zj,\Zj)^{-1}$, and  
%$\Ahi = k(\X_i,\Zhi)k(\Zhi,\Zhi)^{-1}$. \\

@@

Since the optimal distributions $q(\u_j)$ and $q(\v_i)$ are in the exponential family, it is more convenient to use stochastic variational inference \citep{hensman2012fast,hensmangaussian} to perform update of their canonical parameters.
This works by taking a step of length $l$ in the direction of the natural gradient approximated by mini-batches of the data.
For instance, consider $q(\u_j)$ whose canonical parameters are $\Phi_1 = \S_j^{-1} \m_j$ and $\Phi_2 = -\frac{1}{2}\S_j^{-1}$.
Their stochastic update equations at time $t + 1$ are given by:
\begin{align}
\nonumber
& \Phi_{1(t+1)} = \S_{j(t)}^{-1} \m_{j(t)}  \\
 & \quad \qquad +  l \bigg(\sum_{i=1}^P \beta_i w_{ij} \A_j(\oi)^T \ynoj - \S_{j(t)}^{-1} \m_{j(t)} \bigg) \\
& \Phi_{2(t+1)} = -\frac{1}{2} \S_{j(t)}^{-1} + l \bigg(\frac{1}{2} \S_{j(t)}^{-1} - \frac{1}{2} \mat{\Lambda} \bigg),
\end{align}
\normalsize
where $\mat{\Lambda} = \Kjzz^{-1} + \sum_{i=1}^P \beta_i w_{ij}^2 \A_j(\oi)^T \A_j(\oi)$.
%
\subsubsection{Inducing Inputs and Hyper-parameters \label{sec:hyperparameters}}
To learn the hyperparameters, which in this model include the mixing weights, the covariance hyperparameters of the latent processes, and the noise precision of each output, we follow standard practice in GP inference.
For this model this involves taking derivatives of the ELBO and applying standard stochastic gradient descent in alternative steps with the variational parameters, much like a variational EM algorithm. 
The derivatives are given in the supplementary material.

Learning of the inducing inputs, which was not considered in the single output case in \citet{hensmangaussian}, is also possible in our stochastic optimization approach.
In the supplementary material, we show that with the help of vectorized operations, the derivatives of the lower bound wrt the inducing inputs can be computed with a cost \emph{independent of the input dimensionality}.
This makes optimizing the inducing locations a practical option that can be important, for example, in high-dimensional problems where fixing the inducing inputs may hinder the performance of the model.
Indeed, our experiments on a large scale multi-output problem show that automatic learning of the inducing inputs can lead to significant performance gain with little overhead in computation. 
%
\subsection{COMPLEXITY ANALYSIS}
In this section we analyze the complexity of the model and compare it to existing multioutput approaches.
For consistency, we first unify common notations used for all models.
We use $P$ as the number of outputs, $N$ as the number of inputs, $Q$ as the number of shared latent processes, $M_t$ as the \emph{total} number of inducing inputs.
It is worth noting that $M_t = (P + Q) \times M$ in our COGP model, assuming that each sparse process has equal number of inducing points. 
Also, COGP have $P$ additional individual processes, one for each output.

### 2014-03-19 paper/model.tex
minor mods to model
3.2 finished

We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
Note that we reserve subscript $i$ for indexing the outputs and their corresponding individual processes ($i = 1 \hdots P$), $j$ for the latent processes ($j = 1 \hdots Q$), and $n$ for the inputs ($n = 1 \hdots N$).
%
\subsection{Prior Model}
From the definition of the GPs and the conditional independence of the processes given the inducing variables, 
the \emph{prior} of the multi-output model can be written as:
\begin{align}
\label{eq:gu}
p(\g | \u) &= \prod_{j=1}^Q p(\g_j | \u_j) = \prod_{j=1}^Q \Normal(\g_j; \BigMu_j, \tilde{\K}_j )\\
\label{eq:u}
p(\u) &= \prod_{j=1}^Q p(\u_j) = \prod_{j=1}^Q \Normal(\u_j; \vec{0}, k(\Zj, \Zj)) \\
\label{eq:hv}
p(\h | \v) &= \prod_{i=1}^P p(\h_i | \v_i) = \prod_{i=1}^P \Normal(\h_i; \BigMu^h_i, \tilde{\K}^h_i)\\
\label{eq:v}
p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)),
\end{align}
where
\begin{align}
 \BigMu_j &= k(\X,\Zj)k(\Zj,\Zj)^{-1}\u_j \\
\BigMu^h_i &= k(\X,\Zhi)k(\Zhi,\Zhi)^{-1}\v_i \\
\tilde{\K}_j &= k_j(\X,\X) - k(\X,\Zj)k(\Zj,\Zj)^{-1}k(\Zj,\X) \\
\tilde{\K}^h_i &= k^h_i(\X,\X) - k(\X,\Zhi)k(\Zhi,\Zhi)^{-1}k(\Zhi,\X).
\end{align}
In the equations and hereafter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.
Here, equations \ref{eq:u} and \ref{eq:v} follow directly from the properties of GPs, while the expressions for $p(\g|\u)$ and $p(\h|\v)$ (eq. \ref{eq:gu} and \ref{eq:hv}) come from the conditionals of the multivariate Gaussian distributions.
Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the important role of the inducing variables in the model.
%
\subsection{Likelihood Model}
As mentioned above, we assume that  observations for each output 
are  (noisy) linear combinations of the $Q$ latent functions $g_{j}(\vec{x})$ plus
an independent function $h_{i}(\vec{x})$. Hence we have that the likelihood 
with standard iid Gaussian noise is given by:
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P \prod_{n=1}^N \Normal( y_{in} ; \sum_{j=1}^Q w_{ij} g_j(\x_n) + h_i(\x_n), \beta_i^{-1}),
\end{align}
where $w_{ij}$ are the corresponding weights and $\beta_i$ is the precision of each Gaussian.
%which says that $y_{in}$ is a linear combination of the latent functions with weight $w_{ij}$ for the $j$th shared latent process plus a contribution from the individual process $h_i$.
As the latent values $\g$ are specified conditioned on the inducing variables $\u$, this construction implies that each output is a weighted combination of the inducing values.
We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor model \citep{teh-et-al-aistats-05}.
However, doing so is against the purpose of our model which encourages sharing of outputs via the inducing variables.
Furthermore, as we shall see in the next section, explicit representation of these variables is fundamental to scalable inference of the model.

>>>>

We denote the collective variables: $\g = \{\g_j\}$, $\h = \{\h_i \}$, $\u = \{\u_j\}$, $\v = \{\v_i\}$, $\Z = \{\Zj\}$, and $\Z^h = \{\Zhi \}$ where $\g_j = \{g_j(\x_n)\}$, $\h_i = \{h_i(\x_n)\}$. 
Note that we reserve subscript $i$ for indexing the outputs and their corresponding individual processes ($i = 1 \hdots P$), $j$ for the latent processes ($j = 1 \hdots Q$), and $n$ for the inputs ($n = 1 \hdots N$).
%
\subsection{PRIOR MODEL}
From the definition of the GPs and the conditional independence of the processes given the inducing variables, 
the \emph{prior} of the multi-output model can be written as:
\begin{align}
\label{eq:gu}
p(\g | \u) &= \prod_{j=1}^Q p(\g_j | \u_j) = \prod_{j=1}^Q \Normal(\g_j; \BigMu_j, \tilde{\K}_j )\\
\label{eq:u}
p(\u) &= \prod_{j=1}^Q p(\u_j) = \prod_{j=1}^Q \Normal(\u_j; \vec{0}, k(\Zj, \Zj)) \\
\label{eq:hv}
p(\h | \v) &= \prod_{i=1}^P p(\h_i | \v_i) = \prod_{i=1}^P \Normal(\h_i; \BigMu^h_i, \tilde{\K}^h_i)\\
\label{eq:v}
p(\v) &= \prod_{i=1}^P p(\v_i) = \prod_{i=1}^P \Normal(\v_i; \vec{0}, k(\Zhi, \Zhi)),
\end{align}
where the corresponding means and covariances of the Gaussians are given by:
\begin{align}
 \BigMu_j &= k(\X,\Zj)k(\Zj,\Zj)^{-1}\u_j \\
\BigMu^h_i &= k(\X,\Zhi)k(\Zhi,\Zhi)^{-1}\v_i \\
\tilde{\K}_j &= k_j(\X,\X) - k(\X,\Zj)k(\Zj,\Zj)^{-1}k(\Zj,\X) \\
\tilde{\K}^h_i &= k^h_i(\X,\X) - k(\X,\Zhi)k(\Zhi,\Zhi)^{-1}k(\Zhi,\X).
\end{align}
In the equations and hereafter, we omit the subscripts $j,h,i$ from the kernels $k_j(\cdot,\cdot)$ and $k^h_i(\cdot,\cdot)$ when it is clear from the parameters inside the parentheses which covariance function is in action.
Here, equations \ref{eq:u} and \ref{eq:v} follow directly from the properties of GPs, while the expressions for $p(\g|\u)$ and $p(\h|\v)$ (eq. \ref{eq:gu} and \ref{eq:hv}) come from the conditionals of the multivariate Gaussian distributions.
Instead of writing the joint priors $p(\g,\u)$ and $p(\h,\v)$, the above equivalent equations are given to emphasize the important role of the inducing variables in the model.
%
\subsection{LIKELIHOOD MODEL}
As mentioned above, we assume that  observations for each output 
are  (noisy) linear combinations of the $Q$ latent functions $g_{j}(\vec{x})$ plus
an independent function $h_{i}(\vec{x})$. Hence we have that the likelihood 
with standard iid Gaussian noise is given by:
\begin{align}
p(\y | \g, \h ) = \prod_{i=1}^P \prod_{n=1}^N \Normal( y_{in} ; \sum_{j=1}^Q w_{ij} g_j(\x_n) + h_i(\x_n), \beta_i^{-1}),
\end{align}
where $w_{ij}$ are the corresponding weights and $\beta_i$ is the precision of each Gaussian.
%which says that $y_{in}$ is a linear combination of the latent functions with weight $w_{ij}$ for the $j$th shared latent process plus a contribution from the individual process $h_i$.
As the latent values $\g$ are specified conditioned on the inducing variables $\u$, this construction implies that each output is a weighted combination of the inducing values.
We note that if $\u$ and $\v$ are marginalized out, we obtain the semiparametric latent factor model \citep{teh-et-al-aistats-05}.
However, doing so is against the purpose of our model which encourages sharing of outputs via the inducing variables.
Furthermore, as we shall see in the next section, explicit representation of these variables is fundamental to scalable inference of the model.
